---
title: "Watershed-scale floodplain delineation — Parsnip River Watershed Group"
output:
  bookdown::html_vignette2
bibliography: references.bib
link-citations: false
vignette: >
  %\VignetteIndexEntry{Watershed-scale floodplain delineation — Parsnip River Watershed Group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi = 150,
  message = FALSE,
  warning = FALSE
)
library(flooded)
```

This vignette runs the `flooded` pipeline on the Parsnip River Watershed
Group (`PARS`, 5,597 km², north-eastern BC) and demonstrates the
AOI-driven helper [`flooded::fl_dem_aoi()`][fl_dem_aoi].

The Parsnip River Watershed Group sits between Prince George and
Mackenzie, BC. The Parsnip River flows north and enters the southern
arm of Williston Reservoir at Mackenzie, joining the Peace River
system. From there the drainage runs Peace → Slave → Mackenzie River,
ultimately discharging to the Arctic Ocean via the Mackenzie Delta.

[fl_dem_aoi]: ../reference/fl_dem_aoi.html

## Why floodplains

Floodplains are the low-lying, periodically inundated lands flanking a
stream — the parts of the valley that water reclaims when discharge
exceeds bankfull. They store flood water and dissipate flood energy,
recharge shallow groundwater, sort sediment, and host the off-channel
ponds, side channels, sloughs, lakes, and wetlands that drive
disproportionate ecological productivity per unit area. For salmonids
in particular, floodplain habitat — slow-water margins, beaver
complexes, off-channel rearing — is often the limiting factor for
freshwater survival, especially through winter low-flow and high-flow
refugia. Mapping the *extent* of the floodplain at watershed scale is
therefore the first step in scoping where restoration, protection, or
flood-risk planning has the most leverage.

## Modelling parameters

`flooded::fl_valley_confine()` combines four masks — slope, distance
from stream, cost-distance through the terrain, and a bankfull flood
model — and applies morphological cleanup (closing, hole-fill, patch
removal) before returning a binary valley raster. The algorithm is
adapted from the
[USDA Valley Confinement Algorithm](https://github.com/Ecotrust/BlueGeo)
[@nagel_etal2014LandscapeScale] (BlueGeo R port by Devin Cairns, MIT)
and the [bcfishpass](https://github.com/smnorris/bcfishpass) lateral
habitat assembly (Simon Norris, Apache 2.0). The bankfull flood model
follows the regional regression of @hall_etal2007Predictingriver. The
parameter legend (units, defaults, citations, ecological effect) lives
in `flooded::fl_params()`.

This run uses the **`ff04`** scenario from `flooded::fl_scenarios()` —
*functional floodplain*, the recurrent-inundation footprint. The active
parameter values are shown in the table below.

```{r citation-helper, include = FALSE}
# Convert "key1;key2" bare-key strings (the format used in
# `fl_params()$citation_keys` and `fl_scenarios()$citation_keys`) into
# pandoc citation markers `[@key1; @key2]`. The output is then handed to
# `xciter::xct_keys_to_inline_table_col()` which calls pandoc to render
# them as proper inline citations, against the same references.bib that
# pandoc uses for the rest of the vignette.
keys_to_pandoc <- function(s) {
  vapply(s, function(x) {
    if (is.na(x) || !nzchar(x)) return("")
    paste0("[", paste0("@", trimws(strsplit(x, ";")[[1]]), collapse = "; "), "]")
  }, character(1), USE.NAMES = FALSE)
}
```

```{r params-stamp, echo = FALSE}
ff_used <- 4
pars_used <- c("flood_factor", "slope_threshold", "max_width", "cost_threshold")
p_tab <- flooded::fl_params()[match(pars_used, flooded::fl_params()$parameter),
                              c("parameter", "default", "unit", "effect",
                                "citation_keys")]
p_tab$value <- p_tab$default
p_tab$value[p_tab$parameter == "flood_factor"] <- ff_used
p_tab$citation_keys <- keys_to_pandoc(p_tab$citation_keys)
p_tab <- xciter::xct_keys_to_inline_table_col(
  p_tab, col_format = "citation_keys",
  path_bib = "references.bib"
)
p_tab <- p_tab[, c("parameter", "value", "unit", "effect", "citation_keys")]
names(p_tab)[names(p_tab) == "citation_keys"] <- "source"
knitr::kable(
  p_tab, row.names = FALSE,
  caption = "Active parameter values for the Parsnip River Watershed Group run. `flood_factor = 4` (`ff04`); other values are package defaults from `flooded::fl_params()`."
)
```

The package ships three pre-baked scenarios (table below). They differ
only in `flood_factor`; all other parameters are held constant so
output differences isolate the ecological signal.

```{r scenarios-stamp, echo = FALSE}
s_tab <- flooded::fl_scenarios()[, c("scenario_id", "flood_factor",
                                     "description", "ecological_process",
                                     "citation_keys")]
s_tab$citation_keys <- keys_to_pandoc(s_tab$citation_keys)
s_tab <- xciter::xct_keys_to_inline_table_col(
  s_tab, col_format = "citation_keys",
  path_bib = "references.bib"
)
names(s_tab)[names(s_tab) == "citation_keys"] <- "source"
knitr::kable(
  s_tab, row.names = FALSE,
  caption = "Pre-baked flood-factor scenarios shipped with the package. Switch by passing `flood_factor =` to `flooded::fl_valley_confine()`."
)
```

For a side-by-side comparison of all three scenarios on a smaller reach,
see the
[valley-confinement vignette](https://newgraphenvironment.github.io/flooded/articles/valley-confinement.html).

## Cached inputs

The vignette renders against pre-computed outputs from
`data-raw/wsg_vignette_data.R` so it builds fast and doesn't touch the
network or database. The hydro / context layers ship as a single
multi-layer GeoPackage and the rasters as separate GeoTIFFs (raster
tiles in GPKG would lose the continuous DTM precision and the binary
valleys semantics — wrong format for analytical layers).

Direct downloads of the cached Parsnip River Watershed Group bundle from the repo (open in QGIS
or any GDAL-aware tool):

- [`pars.gpkg`](https://github.com/NewGraphEnvironment/flooded/raw/main/inst/vignette-data/pars.gpkg)
  — vectors: `aoi`, `streams`, `waterbodies`, `floodplain`, `railways`,
  `roads`, `reserves`, `parks`, `named_streams`
- [`pars_dem.tif`](https://github.com/NewGraphEnvironment/flooded/raw/main/inst/vignette-data/pars_dem.tif)
  — MRDEM-30 clip (Int16)
- [`pars_valleys.tif`](https://github.com/NewGraphEnvironment/flooded/raw/main/inst/vignette-data/pars_valleys.tif)
  — `fl_valley_confine()` binary output

```{r load, include = FALSE}
library(terra)
library(sf)

gpkg    <- system.file("vignette-data/pars.gpkg",   package = "flooded", mustWork = TRUE)
dem_tif <- system.file("vignette-data/pars_dem.tif", package = "flooded", mustWork = TRUE)
val_tif <- system.file("vignette-data/pars_valleys.tif", package = "flooded", mustWork = TRUE)

present <- sf::st_layers(gpkg)$name
load_layer <- function(name) {
  if (name %in% present) sf::st_read(gpkg, layer = name, quiet = TRUE) else NULL
}

aoi           <- load_layer("aoi")
streams       <- load_layer("streams")
waterbodies   <- load_layer("waterbodies")
floodplain    <- load_layer("floodplain")
railways      <- load_layer("railways")
roads         <- load_layer("roads")
reserves      <- load_layer("reserves")
parks         <- load_layer("parks")
named_streams <- load_layer("named_streams")
dem     <- terra::rast(dem_tif)
valleys <- terra::rast(val_tif)

# Trim the DEM to the WSG and clip every layer to the same boundary so
# nothing renders outside the basin.
dem        <- terra::mask(terra::crop(dem, terra::vect(aoi)), terra::vect(aoi))
valleys    <- terra::mask(terra::crop(valleys, terra::vect(aoi)), terra::vect(aoi))
streams    <- fresh::frs_clip(streams, aoi)
waterbodies <- fresh::frs_clip(waterbodies, aoi)
floodplain <- fresh::frs_clip(floodplain, aoi)
if (!is.null(roads))         roads         <- fresh::frs_clip(roads, aoi)
if (!is.null(railways))      railways      <- fresh::frs_clip(railways, aoi)
if (!is.null(reserves))      reserves      <- fresh::frs_clip(reserves, aoi)
if (!is.null(parks))         parks         <- fresh::frs_clip(parks, aoi)
if (!is.null(named_streams)) named_streams <- fresh::frs_clip(named_streams, aoi)
```

## DEM

The underlying elevation data is
[MRDEM-30](https://open.canada.ca/data/en/dataset/18752265-bda3-498c-a4ba-9dfe68cb98da),
NRCan's 30 m Medium-Resolution Digital Elevation Model — a single
Cloud-Optimized GeoTIFF covering all of Canada, hosted on public S3
with no authentication required, integrated from LidarBC where lidar
coverage exists with Copernicus TanDEM-X / CDEM-derived fallback
elsewhere. `flooded::fl_dem_aoi()` reads it via `/vsicurl/`, so the only
bytes transferred are those intersecting the AOI — bandwidth scales
with the AOI, not with the COG size.

The DEM was fetched in `data-raw/wsg_vignette_data.R` with a single
call, passing the streams as the AOI so the crop hugs the stream
corridor (a memory-efficient pattern for large watersheds with sparse
stream networks):

```{r dem-fetch, eval = FALSE}
dem <- flooded::fl_dem_aoi(streams, buffer = 2000)
```

The default `source = NULL` resolves to the canonical MRDEM-30 DTM
`/vsicurl/` URL inside `flooded::fl_dem_aoi()`. `buffer = 2000` extends
the AOI by 2 km in metres before crop. To override the source — e.g.,
to fetch a LidarBC COG — pass `source = "/vsicurl/https://.../tile.tif"`.

## Streams and waterbodies

Streams are habitat segments modelled as accessible to bull trout from
[bcfishpass](https://github.com/smnorris/bcfishpass) outputs. bcfishpass
runs weekly on a hosted virtual machine and republishes the
`streams_bt_vw` view, so the data is current to within a week of any
upstream FWA, observation, or barrier change.

```{r bcfishpass-version, echo = FALSE}
ver <- readRDS(system.file("vignette-data/pars_meta.rds",
                           package = "flooded", mustWork = TRUE))
```

This run is built against bcfishpass `r ver$bcfishpass_version` (LINEAR
model, completed `r ver$bcfishpass_date`). The version stamp is cached
at data-raw time (see `data-raw/wsg_vignette_data.R`) so the vignette
renders without touching the database.

The streams query is a single
[`fresh::frs_db_query()`](https://newgraphenvironment.github.io/fresh/reference/frs_db_query.html)
call — every row in `bcfishpass.streams_bt_vw` is already classified
accessible (`access IN (1, 2)` = assessed or modelled), so filtering to
"best accessible habitat order 3+" needs no working-table classification:

```r
streams <- fresh::frs_db_query(conn, "
  SELECT segmented_stream_id, blue_line_key, waterbody_key,
         upstream_area_ha, map_upstream, channel_width, stream_order,
         gradient, mapping_code, access, spawning, rearing, geom
  FROM bcfishpass.streams_bt_vw
  WHERE watershed_group_code = 'PARS'
    AND access IN (1, 2)
    AND stream_order >= 3")
```

Waterbodies (lakes + wetlands) come from `whse_basemapping.fwa_lakes_poly`
and `whse_basemapping.fwa_wetlands_poly` joined on `waterbody_key` — only
those physically anchored to the streams above are pulled in. They feed
`flooded::fl_valley_confine(waterbodies = ...)` so the final valley
raster fills lake / wetland cells the gradient and cost-distance masks
would otherwise carve donut holes around.

## Valley confinement

`flooded::fl_valley_confine()` runs the full VCA pipeline on the cached
DEM, streams, and waterbodies. On the Parsnip River Watershed Group at 30 m (~20 Mcells) it takes a
couple of minutes single-threaded; `terra::terraOptions(threads = N)`
parallelises the heavy raster ops.

```{r vca, eval = FALSE}
valleys <- flooded::fl_valley_confine(
  dem = dem,
  streams = streams,
  field = "upstream_area_ha",
  precip = flooded::fl_stream_rasterize(streams, dem, field = "map_upstream"),
  waterbodies = waterbodies,
  flood_factor = 4   # ff04 — functional floodplain
)
floodplain <- flooded::fl_valley_poly(valleys)
```

## Floodplain map — full WSG

```{r map-floodplain, echo = FALSE, fig.cap = "Parsnip River Watershed Group unconfined valleys (green) over MRDEM-30 hillshade. Parks (light green polygon), First Nations reserves (light grey polygon, black diamond marker at centroid + label), accessible order 3+ streams in blue, lakes and wetlands in light blue, forest service / resource roads grey, railways black-dashed, watershed boundary heavy black."}
hill <- terra::shade(
  terra::terrain(dem, "slope", unit = "radians"),
  terra::terrain(dem, "aspect", unit = "radians")
)

terra::plot(hill, col = grey(0:100 / 100), legend = FALSE, axes = FALSE,
            mar = c(2, 2, 2, 2),
            main = "Parsnip River Watershed Group — functional floodplain (ff04), accessible order 3+")

# Hydro / floodplain
terra::plot(valleys, col = c(NA, "#2c7a2c"), alpha = 0.55,
            add = TRUE, legend = FALSE)
plot(sf::st_geometry(waterbodies), col = "#a3cdb966", border = "#2171B5",
     lwd = 0.3, add = TRUE)
plot(sf::st_geometry(streams), col = "#1f77b4", lwd = 0.4, add = TRUE)

# Transport — FSRs only on the wide view; full network too cluttered
if (!is.null(roads)) {
  fsr <- roads[roads$transport_line_type_code %in% c("RRS", "RRD", "RRN"), ]
  if (nrow(fsr) > 0L) {
    plot(sf::st_geometry(fsr), col = "#88888888", lwd = 0.25, add = TRUE)
  }
}
if (!is.null(railways)) {
  plot(sf::st_geometry(railways), col = "black", lwd = 0.8, lty = "dashed",
       add = TRUE)
}

# Land-use polygons drawn ABOVE hydro/transport so they're visible where
# they overlap. Colours from the gq registry (`reg_qgis_restoration.json`).
if (!is.null(parks)) {
  plot(sf::st_geometry(parks), col = "#639b5f55", border = "#33a02c",
       lwd = 0.7, add = TRUE)
}
if (!is.null(reserves)) {
  plot(sf::st_geometry(reserves), col = "#b2b2b288", border = "#232323",
       lwd = 0.6, add = TRUE)
  # Reserves are tiny at WSG scale — overlay a point marker at each
  # centroid so they read on the map.
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves),
                                          of_largest_polygon = TRUE))
  plot(pts, pch = 23, bg = "#222222", col = "white", cex = 0.9, add = TRUE)
}

plot(sf::st_geometry(aoi), border = "black", lwd = 1.5, add = TRUE)

# Reserve labels — drawn last and with xpd = TRUE so a label whose
# centroid sits near the AOI edge after clipping isn't cropped at the
# plot region. Tiny font with a thin white halo for legibility on
# hillshade.
if (!is.null(reserves) && "english_name" %in% names(reserves) &&
    nrow(reserves) > 0L) {
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  ofs_y <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) {
    text(coords[, 1] + dx * halo_x, ly, labels = reserves$english_name,
         cex = 0.45, col = "white", font = 2, xpd = TRUE)
  }
  for (dy in c(-1, 1)) {
    text(coords[, 1], ly + dy * halo_y, labels = reserves$english_name,
         cex = 0.45, col = "white", font = 2, xpd = TRUE)
  }
  text(coords[, 1], ly, labels = reserves$english_name,
       cex = 0.45, col = "#111111", font = 2, xpd = TRUE)
}
```

## Detail map — south-east corner

The full-WSG view compresses a lot of detail. Cropping to the
south-eastern corner — the headwaters of the Parsnip River, where Arctic
Lake Provincial Park sits on the continental divide between Arctic
Ocean and Pacific Ocean drainages — shows the per-reach floodplain
pattern, individual channels, lakes / wetlands, and the named First
Nations reserves at full resolution. Streams just inside the Parsnip
River Watershed Group here drain north via Williston Reservoir → Peace
→ Slave → Mackenzie to the Arctic; a short distance south of the
divide, streams drain into the Fraser and out to the Pacific.

```{r map-detail, echo = FALSE, fig.cap = "South-east corner of the Parsnip River Watershed Group at full resolution — the headwaters around Arctic Lake on the continental divide. Parks (light green), First Nations reserves (light grey polygon with black diamond marker + formal english_name label at centroid), waterbodies, valleys, named streams (italic blue labels), roads (grey), railways (black dashed)."}
e <- sf::st_bbox(aoi)
inset_bbox <- sf::st_bbox(c(
  xmin = unname(e["xmin"] + (e["xmax"] - e["xmin"]) * 0.55),
  ymin = unname(e["ymin"]),
  xmax = unname(e["xmax"]),
  ymax = unname(e["ymin"] + (e["ymax"] - e["ymin"]) * 0.45)
), crs = sf::st_crs(streams))
inset_ext <- terra::ext(inset_bbox[c("xmin", "xmax", "ymin", "ymax")])

crop_sf <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(NULL)
  out <- suppressWarnings(sf::st_crop(x, inset_bbox))
  if (nrow(out) == 0L) NULL else out
}

dem_inset            <- terra::crop(dem, inset_ext)
valleys_inset        <- terra::crop(valleys, inset_ext)
streams_inset        <- crop_sf(streams)
wb_inset             <- crop_sf(waterbodies)
aoi_inset            <- crop_sf(aoi)
roads_inset          <- crop_sf(roads)
railways_inset       <- crop_sf(railways)
reserves_inset       <- crop_sf(reserves)
parks_inset          <- crop_sf(parks)
named_streams_inset  <- crop_sf(named_streams)

hill_inset <- terra::shade(
  terra::terrain(dem_inset, "slope", unit = "radians"),
  terra::terrain(dem_inset, "aspect", unit = "radians")
)

terra::plot(hill_inset, col = grey(0:100 / 100), legend = FALSE, axes = FALSE,
            mar = c(2, 2, 2, 2),
            main = "Parsnip River Watershed Group — south-east detail (continental divide)")

# Hydro / floodplain
terra::plot(valleys_inset, col = c(NA, "#2c7a2c"), alpha = 0.55,
            add = TRUE, legend = FALSE)
if (!is.null(wb_inset)) {
  plot(sf::st_geometry(wb_inset), col = "#a3cdb988", border = "#2171B5",
       lwd = 0.4, add = TRUE)
}
plot(sf::st_geometry(streams_inset), col = "#1f77b4", lwd = 0.7, add = TRUE)

# Transport
if (!is.null(roads_inset)) {
  plot(sf::st_geometry(roads_inset), col = "#666666aa", lwd = 0.4, add = TRUE)
}
if (!is.null(railways_inset)) {
  plot(sf::st_geometry(railways_inset), col = "black", lwd = 1.0, lty = "dashed",
       add = TRUE)
}

# Land-use polygons on top so they read on overlap.
if (!is.null(parks_inset)) {
  plot(sf::st_geometry(parks_inset), col = "#639b5f55", border = "#33a02c",
       lwd = 0.8, add = TRUE)
}
if (!is.null(reserves_inset)) {
  plot(sf::st_geometry(reserves_inset), col = "#b2b2b2aa", border = "#232323",
       lwd = 0.7, add = TRUE)
}

if (!is.null(aoi_inset)) {
  plot(sf::st_geometry(aoi_inset), border = "black", lwd = 1.5, add = TRUE)
}

# Named-stream labels — one per unique gnis_name at the segment midpoint
if (!is.null(named_streams_inset)) {
  dedup <- named_streams_inset[!duplicated(named_streams_inset$gnis_name), ]
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(dedup)))
  text(sf::st_coordinates(pts), labels = dedup$gnis_name,
       cex = 0.5, font = 3, col = "#0d3a6c")
}

# First Nations reserve markers + labels (inset only)
if (!is.null(reserves_inset) && "english_name" %in% names(reserves_inset) &&
    nrow(reserves_inset) > 0L) {
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves_inset),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  plot(pts, pch = 23, bg = "#222222", col = "white", cex = 1.0, add = TRUE)
  ofs_y <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) {
    text(coords[, 1] + dx * halo_x, ly, labels = reserves_inset$english_name,
         cex = 0.45, col = "white", font = 2, xpd = TRUE)
  }
  for (dy in c(-1, 1)) {
    text(coords[, 1], ly + dy * halo_y, labels = reserves_inset$english_name,
         cex = 0.45, col = "white", font = 2, xpd = TRUE)
  }
  text(coords[, 1], ly, labels = reserves_inset$english_name,
       cex = 0.45, col = "#111111", font = 2, xpd = TRUE)
}
```

## References
