---
title: "Watershed-scale floodplain delineation — PARS"
output:
  bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Watershed-scale floodplain delineation — PARS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi = 150,
  message = FALSE,
  warning = FALSE
)
```

This vignette runs the `flooded` pipeline on a full BC watershed group: the
**Parsnip River** (`PARS`), 5,597 km² in the FWCP Peace region. It's the
reference example for the FWCP Peace floodplain reporting work, and it
demonstrates the AOI-driven helper [`fl_dem_aoi()`][fl_dem_aoi] that pulls a
DEM straight from
[NRCan's MRDEM-30](https://open.canada.ca/data/en/dataset/18752265-bda3-498c-a4ba-9dfe68cb98da)
without any per-project DEM plumbing.

[fl_dem_aoi]: ../reference/fl_dem_aoi.html

The vignette loads pre-computed outputs from `inst/vignette-data/`. The
underlying data was generated by `data-raw/wsg_vignette_data.R`, which is
generic — change `wsg <- "PARS"` to any 4-letter BC watershed group code
and re-run to swap in another basin (the BULK, KOTL, ATNA, etc. cases drop
straight in).

## Why MRDEM-30

The DEM choice is settled in
[`rtj/docs/dem-sources.md`](https://github.com/NewGraphEnvironment/rtj/blob/main/docs/dem-sources.md).
For watershed-scale floodplain work, MRDEM-30 is the primary path:

- 30 m, all of Canada, Cloud-Optimized GeoTIFF on public S3 (no auth)
- Integrates LidarBC where available; falls back to Copernicus / TanDEM-X
  for areas without lidar (e.g., parts of the Peace)
- Single 84 GB COG accessed via `/vsicurl/` — `terra::rast()` reads only
  the bytes intersecting an AOI, so bandwidth scales with AOI size, not
  COG size

For sub-10 m riparian-scale work where lidar coverage exists, the
`stac-dem-bc` vignette covers the LidarBC STAC pattern; `fl_dem_aoi(source = ...)`
accepts those COG URLs directly.

## Load data

```{r load}
library(flooded)
library(terra)
library(sf)

vd <- function(name) {
  system.file("vignette-data", name, package = "flooded", mustWork = TRUE)
}

aoi        <- st_read(vd("pars_aoi.gpkg"), quiet = TRUE)
streams    <- st_read(vd("pars_streams.gpkg"), quiet = TRUE)
dem        <- rast(vd("pars_dem.tif"))
valleys    <- rast(vd("pars_valleys.tif"))
floodplain <- st_read(vd("pars_floodplain.gpkg"), quiet = TRUE)
```

## Streams

PARS streams come from `bcfishpass.streams_vw`, filtered to Strahler order
4+. The view carries the attributes the flood model needs
(`channel_width`, `upstream_area_ha`, `map_upstream`) — the bare
`whse_basemapping.fwa_stream_networks_sp` doesn't.

```{r streams-summary}
nrow(streams)
range(streams$stream_order)
```

## DEM

The DEM was fetched in `data-raw/wsg_vignette_data.R` with a single call,
passing the streams as the AOI so the crop hugs the stream corridor (a
memory-efficient pattern for large watersheds with sparse stream networks):

```{r dem-fetch, eval = FALSE}
# data-raw/wsg_vignette_data.R
dem <- fl_dem_aoi(streams, buffer = 2000)
```

The default `source = NULL` resolves to the canonical MRDEM-30 DTM
`/vsicurl/` URL inside `fl_dem_aoi()`. `buffer = 2000` extends the AOI by
2 km in metres before crop. To override the source — e.g., to fetch a
LidarBC COG — pass `source = "/vsicurl/https://.../tile.tif"`.

```{r dem-summary}
ext(dem)
res(dem)
crs(dem, describe = TRUE)[, "code"]
```

## Valley confinement

`fl_valley_confine()` runs the full VCA pipeline: slope mask, distance
mask, cost-distance mask, and bankfull flood model, with morphological
cleanup. On PARS at 30 m (~20 Mcells) it takes a couple of minutes
single-threaded; `terra::terraOptions(threads = N)` parallelises the
heavy raster ops.

```{r vca, eval = FALSE}
# data-raw/wsg_vignette_data.R
valleys <- fl_valley_confine(
  dem = dem,
  streams = streams,
  field = "upstream_area_ha",
  precip = fl_stream_rasterize(streams, dem, field = "map_upstream")
)
floodplain <- fl_valley_poly(valleys)
```

## Floodplain map

```{r map-floodplain, fig.cap = "PARS unconfined valleys (green) over MRDEM-30 hillshade. Streams in blue, watershed boundary in black."}
hill <- shade(
  terrain(dem, "slope", unit = "radians"),
  terrain(dem, "aspect", unit = "radians")
)

plot(hill, col = grey(0:100 / 100), legend = FALSE, axes = FALSE,
     mar = c(2, 2, 2, 2), main = "PARS — order 4+ floodplain")
plot(valleys, col = c(NA, "#2c7a2c"), alpha = 0.55, add = TRUE,
     legend = FALSE)
plot(st_geometry(streams), col = "#1f77b4", lwd = 0.4, add = TRUE)
plot(st_geometry(aoi), border = "black", lwd = 1.5, add = TRUE)
```

## Summary

```{r summary}
floodplain_area_km2 <- sum(values(valleys) == 1L, na.rm = TRUE) *
  prod(res(valleys)) / 1e6
wsg_area_km2 <- as.numeric(aoi$area_ha) / 100
data.frame(
  watershed_group        = aoi$watershed_group_code,
  wsg_area_km2           = round(wsg_area_km2, 0),
  floodplain_area_km2    = round(floodplain_area_km2, 1),
  floodplain_pct_of_wsg  = round(100 * floodplain_area_km2 / wsg_area_km2, 1),
  n_streams              = nrow(streams),
  n_floodplain_polygons  = nrow(floodplain)
)
```

## Adapting to another watershed group

Open `data-raw/wsg_vignette_data.R`, change `wsg <- "PARS"` to any
4-letter BC WSG code, and re-run. The script writes outputs prefixed with
the lowercase WSG code (e.g., `bulk_*` instead of `pars_*`), so multiple
WSG caches can coexist in `inst/vignette-data/`. Point the vignette's
`vd()` helper at the new prefix and rebuild.

This vignette is also designed to drop into a bookdown report appendix —
swap the YAML output for the report's bookdown config and the chunks
work as-is.
