---
title: "Bring your own DEM via STAC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring your own DEM via STAC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi = 150
)
```

`flooded` works with any DEM — it doesn't require a specific source. This
vignette demonstrates fetching 1 m lidar tiles from a STAC catalog and
comparing the resulting floodplain with the bundled 10 m DEM.

We use the same Neexdzii Kwah test area (Bulkley River near Topley, BC) as
the main vignette, but source the DEM from
[stac-dem-bc](https://images.a11s.one/) — a STAC collection of BC provincial
lidar DEMs.

## Setup

```{r load}
library(flooded)
library(terra)
library(sf)
library(rstac)
library(gdalcubes)

terra::terraOptions(threads = 12)
```

## Bundled 10 m baseline

First, load the bundled test data and run the VCA to establish a baseline.

```{r baseline}
dem_10m <- rast(system.file("testdata/dem.tif", package = "flooded"))
slope_10m <- rast(system.file("testdata/slope.tif", package = "flooded"))
streams <- st_read(
  system.file("testdata/streams.gpkg", package = "flooded"),
  quiet = TRUE
)

precip_r <- fl_stream_rasterize(streams, dem_10m, field = "map_upstream")

valleys_10m <- fl_valley_confine(
  dem_10m, streams,
  field = "upstream_area_ha",
  slope = slope_10m,
  slope_threshold = 9,
  max_width = 2000,
  cost_threshold = 2500,
  flood_factor = 6,
  precip = precip_r
)

n_10m <- sum(values(valleys_10m) == 1, na.rm = TRUE)
cat("10 m DEM valley cells:", n_10m, "/", ncell(valleys_10m),
    "(", round(100 * n_10m / ncell(valleys_10m), 1), "%)\n")
```

## Fetch 1 m lidar DEM from STAC

Query the `stac-dem-bc` collection for tiles intersecting our test area.
We filter to 2019 tiles only (1 m lidar).

```{r stac-query}
# Test area bounding box in WGS84
# terra ext() gives xmin,xmax,ymin,ymax — STAC needs xmin,ymin,xmax,ymax
dem_wgs <- project(dem_10m, "EPSG:4326")
e <- ext(dem_wgs)
bbox <- c(e$xmin, e$ymin, e$xmax, e$ymax)

# Query STAC
items <- stac("https://images.a11s.one/") |>
  stac_search(
    collections = "stac-dem-bc",
    bbox = bbox,
    datetime = "2019-01-01T00:00:00Z/2019-12-31T23:59:59Z"
  ) |>
  post_request() |>
  items_fetch()

cat("STAC items found:", length(items$features), "\n")
for (f in items$features) cat(" ", f$id, "\n")
```

## Mosaic and crop with gdalcubes

Use `gdalcubes` to mosaic the STAC tiles and reproject to BC Albers
(EPSG:3005), matching the test area extent. We use 5 m resolution as a
practical compromise — finer than the bundled 10 m, but fast enough to build
in a vignette. For production work, set `dx = 1, dy = 1` for full 1 m
resolution.

```{r mosaic}
# Build gdalcubes image collection from STAC items
col <- stac_image_collection(items$features, asset_names = "image")

# Define target cube — same extent as bundled DEM, 5 m resolution
e <- ext(dem_10m)
stac_res <- 5  # use 1 for full resolution (100x more cells, ~25 min fetch)
v <- cube_view(
  srs = "EPSG:3005",
  extent = list(
    left = e$xmin, right = e$xmax,
    bottom = e$ymin, top = e$ymax,
    t0 = "2019-01-01", t1 = "2019-12-31"
  ),
  dx = stac_res, dy = stac_res, dt = "P1Y",
  aggregation = "first",
  resampling = "bilinear"
)

# write_tif creates a directory with timestamped TIF(s) inside
cube <- raster_cube(col, v)
out_dir <- tempfile()
write_tif(cube, out_dir)
dem_stac_path <- list.files(out_dir, pattern = "\\.tif$", full.names = TRUE)[1]
dem_stac <- rast(dem_stac_path)

cat(stac_res, "m DEM:", ncol(dem_stac), "x", nrow(dem_stac), "pixels\n")
cat("Resolution:", res(dem_stac), "m\n")
```

## Derive slope

```{r slope}
slope_stac <- terra::terrain(dem_stac, v = "slope", unit = "degrees")
# Convert to percent slope to match flooded convention
slope_stac <- tan(slope_stac * pi / 180) * 100
```

## Run VCA on STAC DEM

```{r vca-stac}
# Rasterize streams onto the STAC DEM grid
precip_stac <- fl_stream_rasterize(streams, dem_stac, field = "map_upstream")

valleys_stac <- fl_valley_confine(
  dem_stac, streams,
  field = "upstream_area_ha",
  slope = slope_stac,
  slope_threshold = 9,
  max_width = 2000,
  cost_threshold = 2500,
  flood_factor = 6,
  precip = precip_stac
)

n_stac <- sum(values(valleys_stac) == 1, na.rm = TRUE)
cat(stac_res, "m DEM valley cells:", n_stac, "/", ncell(valleys_stac),
    "(", round(100 * n_stac / ncell(valleys_stac), 1), "%)\n")
```

## Compare

```{r compare}
# Resample STAC result to 10 m grid for visual comparison
valleys_stac_10 <- resample(valleys_stac, dem_10m, method = "near")

# Area comparison
area_10m <- n_10m * res(dem_10m)[1] * res(dem_10m)[2] / 1e6
area_stac <- sum(values(valleys_stac_10) == 1, na.rm = TRUE) *
  res(dem_10m)[1] * res(dem_10m)[2] / 1e6

cat("Valley area (bundled 10 m DEM):", round(area_10m, 2), "km²\n")
cat("Valley area (STAC", stac_res, "m DEM):", round(area_stac, 2), "km²\n")
```

```{r plot-compare, fig.width = 10, fig.height = 5, fig.cap = "Valley delineation from bundled 10 m DEM (left) vs STAC lidar DEM (right)."}
par(mfrow = c(1, 2))
plot(valleys_10m, col = c("grey90", "darkgreen"),
     main = "Bundled 10 m DEM", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1)

plot(valleys_stac_10, col = c("grey90", "darkgreen"),
     main = paste0("STAC lidar ", stac_res, " m DEM"), legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1)
```

## Summary

Higher resolution DEMs capture finer terrain features — narrow terraces,
incised channels, and small benches that a 10 m DEM smooths over. The
tradeoff is compute time: 1 m grids have 100x more cells than 10 m grids for
the same extent.

The `flooded` pipeline is DEM-agnostic. Any source works:

| Source | Resolution | Access |
|--------|-----------|--------|
| Bundled test data | 10 m | `system.file("testdata/", package = "flooded")` |
| stac-dem-bc (this vignette) | 1–5 m | `rstac` + `gdalcubes` |
| BC Data Catalogue (WCS) | 25 m | `bcdata get-dem` (Python CLI) |
| bcfishpass habitat_lateral | 10 m | Pre-assembled per watershed group |
