---
title: "Valley confinement on Neexdzii Kwah"
output:
  bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Valley confinement on Neexdzii Kwah}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi = 150
)
```

This vignette walks through the full `flooded` pipeline on a section of
Neexdzii Kwah (Bulkley River) near Topley, BC. The test data covers ~8 km of
mainstem plus tributaries (Richfield Creek, Cesford Creek, Robert Hatch Creek)
at 10 m resolution.

## Load data

```{r load}
library(flooded)
library(terra)
library(sf)

dem <- rast(system.file("testdata/dem.tif", package = "flooded"))
slope <- rast(system.file("testdata/slope.tif", package = "flooded"))
streams <- st_read(
  system.file("testdata/streams.gpkg", package = "flooded"),
  quiet = TRUE
)

cat("DEM:", ncol(dem), "x", nrow(dem), "pixels at", res(dem)[1], "m\n")
cat("Streams:", nrow(streams), "segments\n")
cat("Upstream area range:", range(streams$upstream_area_ha), "ha\n")
cat("Mean annual precip range:", range(streams$map_upstream), "mm\n")
```

```{r plot-dem, fig.cap = "DEM with stream network overlay."}
plot(dem, main = "Elevation (m)")
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.5)
```

## Step 1: Rasterize streams

Burn the stream network onto the DEM grid (Figure \@ref(fig:plot-dem)) using
upstream contributing area as the cell value (Figure \@ref(fig:plot-streams)).
This is what the VCA bankfull regression expects — it estimates flood depth
from contributing area in hectares.

```{r rasterize}
stream_r <- fl_stream_rasterize(streams, dem, field = "upstream_area_ha")
```

```{r plot-streams, fig.cap = "Rasterized streams coloured by upstream area (ha)."}
plot(stream_r, main = "Upstream area (ha)")
```

## Step 2: Precipitation

The VCA bankfull regression has a precipitation term that strongly controls
flood depth:

```
bankfull_width = (upstream_area ^ 0.280) * 0.196 * (precip ^ 0.355)
bankfull_depth = bankfull_width ^ 0.607 * 0.145
flood_depth    = bankfull_depth * flood_factor
```

With `precip = 1` (the default), the precipitation term drops out and flood
depths are dramatically underestimated. For the Bulkley mainstem
(`upstream_area_ha ~ 110,000`), the difference is ~2 m vs ~8 m flood depth.

The test data includes `map_upstream` — mean annual precipitation (mm) from
fwapg/ClimateBC, carried as a stream attribute. We rasterize it alongside
the streams to create a spatially varying precipitation surface at stream
cells.

```{r precip}
precip_r <- fl_stream_rasterize(streams, dem, field = "map_upstream")
cat("Precip at stream cells:", range(values(precip_r), na.rm = TRUE), "mm\n")
```

## Step 3: Individual mask components

The VCA combines four spatial criteria. Let's inspect each one.

### Slope mask

Cells with slope <= 9% (the VCA default) represent potentially flat valley
floor (Figure \@ref(fig:plot-slope-mask)).

```{r slope-mask}
mask_slope <- fl_mask(slope, threshold = 9, operator = "<=")
cat("Gentle slope:", sum(values(mask_slope) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-slope-mask, fig.cap = "Slope mask: green = slope <= 9%."}
plot(mask_slope, col = c("grey90", "darkgreen"), main = "Slope <= 9%",
     legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Distance mask

Cells within 1000 m of a stream (half the default `max_width = 2000`;
Figure \@ref(fig:plot-dist-mask)).

```{r distance-mask}
mask_dist <- fl_mask_distance(stream_r, threshold = 1000)
cat("Within 1 km:", sum(values(mask_dist) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-dist-mask, fig.cap = "Distance mask: corridor within 1 km of streams."}
plot(mask_dist, col = c("grey90", "steelblue"), main = "Within 1 km of stream",
     legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Cost distance

Accumulated cost (slope-weighted) distance from streams (Figure \@ref(fig:plot-cost)).
The VCA default threshold is 2500.

```{r cost-distance}
cost <- fl_cost_distance(slope, stream_r)
mask_cost <- fl_mask(cost, threshold = 2500, operator = "<")
cat("Cost < 2500:", sum(values(mask_cost) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-cost, fig.cap = "Cost distance from streams. Warm colours = high cost."}
plot(cost, main = "Cost distance", range = c(0, 5000))
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Flood model

The bankfull regression estimates flood depth from upstream contributing area
and precipitation, interpolates the water surface outward from streams, and
identifies cells below the flood surface (Figure \@ref(fig:plot-flood)).

```{r flood-model}
flood <- fl_flood_model(dem, stream_r, flood_factor = 6, precip = precip_r,
                        max_width = 2000)
cat("Flooded cells:", sum(values(flood[["flooded"]]) == 1, na.rm = TRUE), "\n")
```

```{r plot-flood, fig.cap = "Flood model: depth above terrain (m)."}
depth <- flood[["flood_depth"]]
depth[depth == 0] <- NA
plot(depth, main = "Flood depth (m)")
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

## Step 4: Full VCA pipeline

`fl_valley_confine()` chains all the above (slope + distance + cost + flood
masks), applies morphological cleanup (closing, hole filling, small patch
removal, majority filter), and returns a binary valley raster
(Figure \@ref(fig:plot-valleys)).

Note the `precip` argument — without it, flood depths are ~4x too shallow
and the resulting valley is significantly narrower.

```{r vca}
valleys <- fl_valley_confine(
  dem, streams,
  field = "upstream_area_ha",
  slope_threshold = 9,
  max_width = 2000,
  cost_threshold = 2500,
  flood_factor = 6,
  precip = precip_r
)

n_valley <- sum(values(valleys) == 1, na.rm = TRUE)
cat("Valley cells:", n_valley, "/", ncell(valleys),
    "(", round(100 * n_valley / ncell(valleys), 1), "%)\n")
```

```{r plot-valleys, fig.cap = "Unconfined valleys (green) with stream network."}
plot(valleys, col = c("grey90", "darkgreen"),
     main = "Unconfined valleys", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.2)
```

## Step 5: Post-processing

### Connect to streams

Keep only valley patches that touch a stream cell — removes isolated flat
areas disconnected from the network (Figure \@ref(fig:plot-connected)).

```{r patch-conn}
connected <- fl_patch_conn(valleys, stream_r)
cat("Connected valley cells:",
    sum(values(connected) == 1, na.rm = TRUE), "\n")
```

```{r plot-connected, fig.cap = "Valley patches connected to streams."}
plot(connected, col = c("grey90", "darkgreen"),
     main = "Connected valleys", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.2)
```

### Trim with exclusion masks

`fl_flood_trim()` subtracts user-supplied masks. For example, you could trim
by urban areas, steep terrain, railways, or waterbodies. Here we demonstrate
by removing cells on slopes > 5% (a stricter threshold).

```{r trim}
steep_mask <- fl_mask(slope, threshold = 5, operator = ">")
trimmed <- fl_flood_trim(connected, steep_mask)
cat("After trimming steep cells:",
    sum(values(trimmed) == 1, na.rm = TRUE), "\n")
```

### Assemble multiple layers

`fl_flood_assemble()` unions multiple binary rasters. This is useful when
combining outputs from different flood models or data sources.

```{r assemble}
# Example: union the connected valleys with a wider flood-only mask
flooded_mask <- flood[["flooded"]]
flooded_mask <- ifel(is.na(flooded_mask), 0L, flooded_mask)
assembled <- fl_flood_assemble(connected, flooded_mask)
cat("Assembled cells:", sum(values(assembled) == 1, na.rm = TRUE), "\n")
```

## Precipitation matters

The VCA bankfull regression includes a precipitation term (`precip ^ 0.355`)
that scales flood depth with local climate. In BC, `map_upstream` (mean
annual precipitation in mm) is available as a stream attribute from fwapg
(the Freshwater Atlas Postgres layer). For other jurisdictions, any gridded
precipitation product will work — rasterize it to the DEM grid and pass it
as the `precip` argument.

Omitting precipitation (`precip = 1`, the default) underestimates flood depth
by roughly 4x in wet climates (~500 mm MAP). This produces a valley that is
about half the width of the correct result:

```{r precip-comparison}
valleys_no_precip <- fl_valley_confine(
  dem, streams, field = "upstream_area_ha",
  precip = 1
)

n_no <- sum(values(valleys_no_precip) == 1, na.rm = TRUE)
cat("Without precip:", n_no, "cells (",
    round(100 * n_no / ncell(dem), 1), "%)\n")
cat("With precip:   ", n_valley, "cells (",
    round(100 * n_valley / ncell(dem), 1), "%)\n")
```

## Performance

Several `terra` operations inside `fl_valley_confine()` support multi-threading
(focal filters, distance calculations, raster math). Set threads before running:

```r
terra::terraOptions(threads = 12)  # adjust to your machine
```

On an Apple M4 Max (16 cores, 12 performance), this reduced a Neexdzii Kwah
run (~2,700 km², 27M cells, 1165 stream segments) from ~3.5 minutes to ~1 minute.
The remaining time is dominated by `terra::costDist()` and `terra::interpIDW()`,
which are single-threaded.

## Summary

Key tuning parameters:

| Parameter | Default | Effect |
|-----------|---------|--------|
| `slope_threshold` | 9% | Higher = more valley floor included |
| `max_width` | 2000 m | Analysis corridor width |
| `cost_threshold` | 2500 | Higher = valley extends further up hillslopes |
| `flood_factor` | 6 | Higher = deeper flood, more floodplain |
| `precip` | 1 | MAP in mm — critical for realistic flood depth |
| `size_threshold` | 5000 m² | Minimum valley patch area |
| `hole_threshold` | 2500 m² | Maximum hole size to fill |
