---
title: "Valley confinement on Neexdzii Kwah"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Valley confinement on Neexdzii Kwah}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi = 150
)
```

This vignette walks through the full `flooded` pipeline on a section of
Neexdzii Kwah (Bulkley River) near Topley, BC. The test data covers ~8 km of
mainstem plus tributaries (Richfield Creek, Cesford Creek, Robert Hatch Creek)
at 10 m resolution.

## Load data

```{r load}
library(flooded)
library(terra)
library(sf)

dem <- rast(system.file("testdata/dem.tif", package = "flooded"))
slope <- rast(system.file("testdata/slope.tif", package = "flooded"))
streams <- st_read(
  system.file("testdata/streams.gpkg", package = "flooded"),
  quiet = TRUE
)

cat("DEM:", ncol(dem), "x", nrow(dem), "pixels at", res(dem)[1], "m\n")
cat("Streams:", nrow(streams), "segments\n")
```

```{r plot-dem, fig.cap = "DEM with stream network overlay."}
plot(dem, main = "Elevation (m)")
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.5)
```

## Step 1: Rasterize streams

Burn the stream network onto the DEM grid using channel width as the cell
value. Any numeric attribute works — upstream contributing area is the VCA
default, but channel width is what we have in this dataset.

```{r rasterize}
stream_r <- fl_stream_rasterize(streams, dem, field = "channel_width")
```

```{r plot-streams, fig.cap = "Rasterized streams coloured by channel width (m)."}
plot(stream_r, main = "Channel width (m)")
```

## Step 2: Individual mask components

The VCA combines four spatial criteria. Let's inspect each one before
combining them.

### Slope mask

Cells with slope <= 9% (the VCA default) represent potentially flat valley
floor.

```{r slope-mask}
mask_slope <- fl_mask(slope, threshold = 9, operator = "<=")
cat("Gentle slope:", sum(values(mask_slope) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-slope-mask, fig.cap = "Slope mask: green = slope <= 9%."}
plot(mask_slope, col = c("grey90", "darkgreen"), main = "Slope <= 9%",
     legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Distance mask

Cells within 1000 m of a stream (half the default `max_width = 2000`).

```{r distance-mask}
mask_dist <- fl_mask_distance(stream_r, threshold = 1000)
cat("Within 1 km:", sum(values(mask_dist) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-dist-mask, fig.cap = "Distance mask: corridor within 1 km of streams."}
plot(mask_dist, col = c("grey90", "steelblue"), main = "Within 1 km of stream",
     legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Cost distance

Accumulated cost (slope-weighted) distance from streams. The VCA default
threshold is 2500.

```{r cost-distance}
cost <- fl_cost_distance(slope, stream_r)
mask_cost <- fl_mask(cost, threshold = 2500, operator = "<")
cat("Cost < 2500:", sum(values(mask_cost) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-cost, fig.cap = "Cost distance from streams. Warm colours = high cost."}
plot(cost, main = "Cost distance", range = c(0, 5000))
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Flood model

The bankfull regression estimates flood depth from channel width (or
contributing area), interpolates the water surface outward from streams, and
identifies cells below the flood surface.

```{r flood-model}
flood <- fl_flood_model(dem, stream_r, flood_factor = 6, max_width = 2000)
cat("Flooded cells:", sum(values(flood[["flooded"]]) == 1, na.rm = TRUE), "\n")
```

```{r plot-flood, fig.cap = "Flood model: depth above terrain (m)."}
depth <- flood[["flood_depth"]]
depth[depth == 0] <- NA
plot(depth, main = "Flood depth (m)")
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

## Step 3: Full VCA pipeline

`fl_valley_confine()` chains all the above (slope + distance + cost + flood
masks), applies morphological cleanup (closing, hole filling, small patch
removal, majority filter), and returns a binary valley raster.

```{r vca}
valleys <- fl_valley_confine(
  dem, streams,
  slope_threshold = 9,
  max_width = 2000,
  cost_threshold = 2500,
  flood_factor = 6
)

n_valley <- sum(values(valleys) == 1, na.rm = TRUE)
cat("Valley cells:", n_valley, "/", ncell(valleys),
    "(", round(100 * n_valley / ncell(valleys), 1), "%)\n")
```

```{r plot-valleys, fig.cap = "Unconfined valleys (green) with stream network."}
plot(valleys, col = c("grey90", "darkgreen"),
     main = "Unconfined valleys", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.2)
```

## Step 4: Post-processing

### Connect to streams

Keep only valley patches that touch a stream cell — removes isolated flat
areas disconnected from the network.

```{r patch-conn}
connected <- fl_patch_conn(valleys, stream_r)
cat("Connected valley cells:",
    sum(values(connected) == 1, na.rm = TRUE), "\n")
```

```{r plot-connected, fig.cap = "Valley patches connected to streams."}
plot(connected, col = c("grey90", "darkgreen"),
     main = "Connected valleys", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.2)
```

### Trim with exclusion masks

`fl_flood_trim()` subtracts user-supplied masks. For example, you could trim
by urban areas, steep terrain, or waterbodies. Here we demonstrate by
removing cells on slopes > 5% (a stricter threshold).

```{r trim}
steep_mask <- fl_mask(slope, threshold = 5, operator = ">")
trimmed <- fl_flood_trim(connected, steep_mask)
cat("After trimming steep cells:",
    sum(values(trimmed) == 1, na.rm = TRUE), "\n")
```

### Assemble multiple layers

`fl_flood_assemble()` unions multiple binary rasters. This is useful when
combining outputs from different flood models or data sources.

```{r assemble}
# Example: union the connected valleys with a wider flood-only mask
flooded_mask <- flood[["flooded"]]
flooded_mask <- ifel(is.na(flooded_mask), 0L, flooded_mask)
assembled <- fl_flood_assemble(connected, flooded_mask)
cat("Assembled cells:", sum(values(assembled) == 1, na.rm = TRUE), "\n")
```

## Summary

The full pipeline runs in a few seconds on this 800 x 648 (518k cell) test
dataset. For larger areas, `fl_valley_confine()` benefits from cropping the
DEM to a stream corridor first — the `max_width` parameter implicitly
constrains the analysis domain.

Key tuning parameters:

| Parameter | Default | Effect |
|-----------|---------|--------|
| `slope_threshold` | 9% | Higher = more valley floor included |
| `max_width` | 2000 m | Analysis corridor width |
| `cost_threshold` | 2500 | Higher = valley extends further up hillslopes |
| `flood_factor` | 6 | Higher = deeper flood, more floodplain |
| `size_threshold` | 5000 m² | Minimum valley patch area |
| `hole_threshold` | 2500 m² | Maximum hole size to fill |
