---
title: "Valley confinement from DEM and stream network"
output:
  bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Valley confinement from DEM and stream network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi = 150
)
```

This vignette walks through the full `flooded` pipeline on a section of
Neexdzii Kwah (Bulkley River) near Topley, BC. The test data covers ~8 km of
mainstem plus tributaries (Richfield Creek, Cesford Creek, Robert Hatch Creek)
at 10 m resolution.

The bundled stream network is filtered to order 4+ coho potential habitat
(`bcfishpass.streams_co_vw`). This focuses the floodplain delineation on the
mainstem corridor and major tributaries — the streams most relevant to
restoration planning and where investment has the greatest impact on
higher-value salmon habitat. Filtering also constrains the analysis-of-interest
(AOI) for practical downstream applications like orthophoto acquisition and
review, where cost control matters. All watershed tributaries contribute to
floodplain health, but a bring-your-own-DEM tool benefits from demonstrating
the pipeline on a focused corridor. See `data-raw/network_extract.R` for
the extraction script.

## Load data

```{r load}
library(flooded)
library(terra)
library(sf)

dem <- rast(system.file("testdata/dem.tif", package = "flooded"))
slope <- rast(system.file("testdata/slope.tif", package = "flooded"))
streams <- st_read(
  system.file("testdata/streams.gpkg", package = "flooded"),
  quiet = TRUE
)

cat("DEM:", ncol(dem), "x", nrow(dem), "pixels at", res(dem)[1], "m\n")
cat("Streams:", nrow(streams), "segments\n")
cat("Upstream area range:", range(streams$upstream_area_ha), "ha\n")
cat("Mean annual precip range:", range(streams$map_upstream), "mm\n")
```

```{r plot-dem, fig.cap = "DEM with stream network overlay."}
plot(dem, main = "Elevation (m)")
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.5)
```

## Step 1: Rasterize streams

Burn the stream network onto the DEM grid (Figure \@ref(fig:plot-dem)) using
upstream contributing area as the cell value (Figure \@ref(fig:plot-streams)).
This is what the VCA bankfull regression expects — it estimates flood depth
from contributing area in hectares.

```{r rasterize}
stream_r <- fl_stream_rasterize(streams, dem, field = "upstream_area_ha")
```

```{r plot-streams, fig.cap = "Rasterized streams coloured by upstream area (ha)."}
plot(stream_r, main = "Upstream area (ha)")
```

## Step 2: Precipitation

The VCA bankfull regression has a precipitation term that strongly controls
flood depth:

```
bankfull_width = (upstream_area ^ 0.280) * 0.196 * (precip ^ 0.355)
bankfull_depth = bankfull_width ^ 0.607 * 0.145
flood_depth    = bankfull_depth * flood_factor
```

With `precip = 1` (the default), the precipitation term drops out and flood
depths are dramatically underestimated. For the Bulkley mainstem
(`upstream_area_ha ~ 110,000`), the difference is ~2 m vs ~8 m flood depth.

The test data includes `map_upstream` — mean annual precipitation (mm) from
fwapg/ClimateBC, carried as a stream attribute. We rasterize it alongside
the streams to create a spatially varying precipitation surface at stream
cells.

```{r precip}
precip_r <- fl_stream_rasterize(streams, dem, field = "map_upstream")
cat("Precip at stream cells:", range(values(precip_r), na.rm = TRUE), "mm\n")
```

## Step 3: Individual mask components

The VCA combines four spatial criteria. Let's inspect each one.

### Slope mask

Cells with slope <= 9% (the VCA default) represent potentially flat valley
floor (Figure \@ref(fig:plot-slope-mask)).

```{r slope-mask}
mask_slope <- fl_mask(slope, threshold = 9, operator = "<=")
cat("Gentle slope:", sum(values(mask_slope) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-slope-mask, fig.cap = "Slope mask: green = slope <= 9%."}
plot(mask_slope, col = c("grey90", "darkgreen"), main = "Slope <= 9%",
     legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Distance mask

Cells within 1000 m of a stream (half the default `max_width = 2000`;
Figure \@ref(fig:plot-dist-mask)).

```{r distance-mask}
mask_dist <- fl_mask_distance(stream_r, threshold = 1000)
cat("Within 1 km:", sum(values(mask_dist) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-dist-mask, fig.cap = "Distance mask: corridor within 1 km of streams."}
plot(mask_dist, col = c("grey90", "steelblue"), main = "Within 1 km of stream",
     legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Cost distance

Accumulated cost (slope-weighted) distance from streams (Figure \@ref(fig:plot-cost)).
The VCA default threshold is 2500.

```{r cost-distance}
cost <- fl_cost_distance(slope, stream_r)
mask_cost <- fl_mask(cost, threshold = 2500, operator = "<")
cat("Cost < 2500:", sum(values(mask_cost) == 1, na.rm = TRUE), "cells\n")
```

```{r plot-cost, fig.cap = "Cost distance from streams. Warm colours = high cost."}
plot(cost, main = "Cost distance", range = c(0, 5000))
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

### Flood model

The bankfull regression estimates flood depth from upstream contributing area
and precipitation, interpolates the water surface outward from streams, and
identifies cells below the flood surface (Figure \@ref(fig:plot-flood)).

```{r flood-model}
flood <- fl_flood_model(dem, stream_r, flood_factor = 6, precip = precip_r,
                        max_width = 2000)
cat("Flooded cells:", sum(values(flood[["flooded"]]) == 1, na.rm = TRUE), "\n")
```

```{r plot-flood, fig.cap = "Flood model: depth above terrain (m)."}
depth <- flood[["flood_depth"]]
depth[depth == 0] <- NA
plot(depth, main = "Flood depth (m)")
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 0.8)
```

## Step 4: Full VCA pipeline

`fl_valley_confine()` chains all the above (slope + distance + cost + flood
masks), applies morphological cleanup (closing, hole filling, small patch
removal, majority filter), and returns a binary valley raster
(Figure \@ref(fig:plot-valleys)).

Note the `precip` argument — without it, flood depths are ~4x too shallow
and the resulting valley is significantly narrower.

```{r vca}
valleys <- fl_valley_confine(
  dem, streams,
  field = "upstream_area_ha",
  slope_threshold = 9,
  max_width = 2000,
  cost_threshold = 2500,
  flood_factor = 6,
  precip = precip_r
)

n_valley <- sum(values(valleys) == 1, na.rm = TRUE)
cat("Valley cells:", n_valley, "/", ncell(valleys),
    "(", round(100 * n_valley / ncell(valleys), 1), "%)\n")
```

```{r plot-valleys, fig.cap = "Unconfined valleys (green) with stream network."}
plot(valleys, col = c("grey90", "darkgreen"),
     main = "Unconfined valleys", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.2)
```

## Step 5: Post-processing

### Connect to streams

Keep only valley patches that touch a stream cell — removes isolated flat
areas disconnected from the network (Figure \@ref(fig:plot-connected)).

```{r patch-conn}
connected <- fl_patch_conn(valleys, stream_r)
cat("Connected valley cells:",
    sum(values(connected) == 1, na.rm = TRUE), "\n")
```

```{r plot-connected, fig.cap = "Valley patches connected to streams."}
plot(connected, col = c("grey90", "darkgreen"),
     main = "Connected valleys", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1.2)
```

### Trim with exclusion masks

`fl_flood_trim()` subtracts user-supplied masks. For example, you could trim
by urban areas, steep terrain, railways, or waterbodies. Here we demonstrate
by removing cells on slopes > 5% (a stricter threshold).

```{r trim}
steep_mask <- fl_mask(slope, threshold = 5, operator = ">")
trimmed <- fl_flood_trim(connected, steep_mask)
cat("After trimming steep cells:",
    sum(values(trimmed) == 1, na.rm = TRUE), "\n")
```

### Assemble multiple layers

`fl_flood_assemble()` unions multiple binary rasters. This is useful when
combining outputs from different flood models or data sources.

```{r assemble}
# Example: union the connected valleys with a wider flood-only mask
flooded_mask <- flood[["flooded"]]
flooded_mask <- ifel(is.na(flooded_mask), 0L, flooded_mask)
assembled <- fl_flood_assemble(connected, flooded_mask)
cat("Assembled cells:", sum(values(assembled) == 1, na.rm = TRUE), "\n")
```

## Adding waterbodies and channel buffer

The VCA operates on terrain alone — lakes and wetlands appear as donut holes
in the valley raster because the water surface reads differently than
surrounding terrain. Similarly, at coarse DEM resolution the stream channel
itself can be sub-pixel and excluded from the output.

`fl_valley_confine()` addresses both:

- **`channel_buffer`** — auto-detected when streams have a `channel_width`
  column. Buffers each stream segment by half its channel width and adds to
  the output. This is a DEM correction, not a habitat buffer. Streams with
  NA `channel_width` are skipped for the buffer but still included in the
  VCA flood model via the bankfull regression. Users pulling lower-order
  streams may encounter NA widths — the
  [channel width model](https://www.poissonconsulting.ca/f/859859031)
  (Thorley et al., 2021) does not yet cover most first-order streams.
- **`waterbodies`** — an optional `sf` polygon layer of lakes and wetlands.
  Rasterized as-is (no buffer) and added to the output via logical OR.
  No spatial filtering is applied — pre-filter to valley-bottom features
  before calling if headwater waterbodies are not wanted.

```{r load-waterbodies}
waterbodies <- st_read(
  system.file("testdata/waterbodies.gpkg", package = "flooded"),
  quiet = TRUE
)
cat("Waterbodies:", nrow(waterbodies), "features\n")
cat("Types:", paste(names(table(waterbodies$waterbody_type)),
                    table(waterbodies$waterbody_type), sep = "=", collapse = ", "), "\n")
```

```{r vca-features}
valleys_wb <- fl_valley_confine(
  dem, streams,
  field = "upstream_area_ha",
  slope = slope,
  precip = precip_r,
  waterbodies = waterbodies
)

n_wb <- sum(values(valleys_wb) == 1, na.rm = TRUE)
cell_area <- prod(res(dem))
cat("VCA only:          ", round(n_valley * cell_area / 1e4, 1), "ha\n")
cat("VCA + features:    ", round(n_wb * cell_area / 1e4, 1), "ha\n")
cat("Features added:    ", round((n_wb - n_valley) * cell_area / 1e4, 1), "ha\n")
```

```{r plot-features, fig.width = 7, fig.height = 8, fig.cap = "Valley delineation without (top) and with (bottom) waterbodies and channel buffer. Dark green = VCA valley floor, blue lines = streams, orange outlines = waterbody polygons (lakes and wetlands) added via logical OR. Waterbodies fill donut holes left by the terrain-based VCA where flat water surfaces read differently than surrounding floodplain."}
par(mfrow = c(2, 1), mar = c(2, 4, 2, 1))
plot(valleys, col = c("grey90", "darkgreen"),
     main = "VCA only", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1)

plot(valleys_wb, col = c("grey90", "darkgreen"),
     main = "VCA + waterbodies + channel buffer", legend = FALSE)
plot(st_geometry(streams), add = TRUE, col = "blue", lwd = 1)
if (nrow(waterbodies) > 0) {
  plot(st_geometry(waterbodies), add = TRUE, border = "darkorange",
       lwd = 1.2)
}
```

## Precipitation matters

The VCA bankfull regression includes a precipitation term (`precip ^ 0.355`)
that scales flood depth with local climate. In BC, `map_upstream` (mean
annual precipitation in mm) is available as a stream attribute from fwapg
(the Freshwater Atlas Postgres layer). For other jurisdictions, any gridded
precipitation product will work — rasterize it to the DEM grid and pass it
as the `precip` argument.

Omitting precipitation (`precip = 1`, the default) underestimates flood depth
by roughly 4x in wet climates (~500 mm MAP). This produces a valley that is
about half the width of the correct result:

```{r precip-comparison}
valleys_no_precip <- fl_valley_confine(
  dem, streams, field = "upstream_area_ha",
  precip = 1
)

n_no <- sum(values(valleys_no_precip) == 1, na.rm = TRUE)
cat("Without precip:", n_no, "cells (",
    round(100 * n_no / ncell(dem), 1), "%)\n")
cat("With precip:   ", n_valley, "cells (",
    round(100 * n_valley / ncell(dem), 1), "%)\n")
```

## Performance

Several `terra` operations inside `fl_valley_confine()` support multi-threading
(focal filters, distance calculations, raster math). Set threads before running:

```r
terra::terraOptions(threads = 12)  # adjust to your machine
```

On an Apple M4 Max (16 cores, 12 performance), this reduced a Neexdzii Kwah
run (~2,700 km², 27M cells, 1165 stream segments) from ~3.5 minutes to ~1 minute.
The remaining time is dominated by `terra::costDist()` and `terra::interpIDW()`,
which are single-threaded.

## Summary

Key tuning parameters:

| Parameter | Default | Effect |
|-----------|---------|--------|
| `slope_threshold` | 9% | Higher = more valley floor included |
| `max_width` | 2000 m | Analysis corridor width |
| `cost_threshold` | 2500 | Higher = valley extends further up hillslopes |
| `flood_factor` | 6 | Higher = deeper flood, more floodplain |
| `precip` | 1 | MAP in mm — critical for realistic flood depth |
| `waterbodies` | NULL | sf polygons of lakes/wetlands to fill donut holes |
| `channel_buffer` | auto | Buffer streams by channel_width (DEM correction) |
| `size_threshold` | 5000 m² | Minimum valley patch area |
| `hole_threshold` | 2500 m² | Maximum hole size to fill |
